### 2022-03-02 unif/unif.tex
solid

\ourpara{Examples}
We present some examples that demonstrate advantages of our procedure.
The displayed branches of the constructed trees are not necessarily exhaustive.
We abbreviate 
JP-style projection as \textsf{JP\,Proj},
imitation as \textsf{Imit},
identification as \textsf{Id},
\unifrulename{Decompose} as \textsf{Dc}, 
\unifrulename{Dereference} as \textsf{Dr},
$\textsf{Normalize}_\beta$ as $\textsf{N}_\beta$,
and
\textsf{Bind} of a binding $x$ as $\textsf{B(}x\textsf{)}$.
Transitions of the JP procedure are denoted by $\Longrightarrow$.
For the JP transitions we implicitly apply the generated bindings and fully
normalize terms, which significantly shortens JP derivations.

@@

\section{A New Decidable Fragment}
\label{sec:unif:solid-oracle}

>>>>

\ourpara{unif:examples}
We present some examples that demonstrate advantages of our procedure.
The displayed branches of the constructed trees are not necessarily exhaustive.
We abbreviate 
JP-style projection as \textsf{JP\,Proj},
imitation as \textsf{Imit},
identification as \textsf{Id},
\unifrulename{Decompose} as \textsf{Dc}, 
\unifrulename{Dereference} as \textsf{Dr},
$\textsf{Normalize}_\beta$ as $\textsf{N}_\beta$,
and
\textsf{Bind} of a binding $x$ as $\textsf{B(}x\textsf{)}$.
Transitions of the JP procedure are denoted by $\Longrightarrow$.
For the JP transitions we implicitly apply the generated bindings and fully
normalize terms, which significantly shortens JP derivations.

@@

Like the JP procedure, our procedure misses no unifiers:

\begin{theorem}
\label{thm:completeness}
% \looseness=-1
The procedure described in Section \ref{sec:unif:the-unification-procedure}
parametrized by $\mathcal{P}_\mathsf{c}$ is complete, meaning that the
substitutions on the leaves of the constructed tree form a CSU.
More precisely, let $E$ be a multiset of constraints
and let $V$ be the supply of fresh variables provided to the procedure. 
Then for any unifier $\varrho$ of $E$ there exists
a derivation $\left(E,\makeop{id}\right)\longrightarrow^{*}\sigma$ and a
substitution $\theta$ such that for all free variables $X\not\in V$, we have $\varrho X = \theta\sigma X$.
\end{theorem}

Taking a high-level view, this theorem is proved by incrementally defining states
$(E_{j},\sigma_{j})$ and \emph{remainder} \emph{substitutions}
$\varrho_{j}$ starting with
$\left(E_{0},\sigma_{0}\right)=\left(E,\makeop{id}\right)$ and
$\varrho_{0}=\varrho$. The substitution $\varrho_{j}$ is what remains to be added
to $\sigma_{j}$ to reach $\varrho_{0}$. States are defined
so that the shape of the selected constraint from $E_{j}$ and the remainder
substitution guide the choice of applicable transition rule.
We employ a measure based on values of $E_j$ and $\varrho_j$ that decreases with 
each application of the rules. Therefore, eventually, we will reach the target substitution
$\sigma$.

In the remaining of this section, we view terms as
$\alpha\beta\eta$-equivalence classes, with the $\eta$-long $\beta$-normal
form as their canonical representative. Moreover, we consider all
substitutions to be fully applied. These assumptions are justified because
all bindings depend only on the head of terms and hence replacing the lazy
transitions $\unifrulename{Normalize}_{\alpha\eta}$,
$\unifrulename{Normalize}_{\beta}$, and $\unifrulename{Dereference}$ by eager
counterparts only affects the efficiency but not the overall behavior of our
procedure.

We now give the detailed completeness proof of Theorem \ref{thm:completeness}.
Our proof is an adaptation of the proof given by Jensen and Pietrzykowski
\cite{jp-76-unif}.
% %Intimate knowledge of that proof is required from the reader.
Definitions and lemmas are reused, but are combined together differently to
suit our procedure. We start by listing all reused definitions and lemmas
from the original JP proof. The ``JP'' labels in
their statements refer to the corresponding lemmas and definitions from
the original proof.

\begin{defi}[{$\jp[1]D6$}] \label{unif:def:disagreement-pairs}
\looseness=-1
Given two terms $t$ and $s$ and their common context $C$, 
we can write $t$ as $C[\seq t]$ and $s$ as $C[\seq s]$
for some $\seq t$ and $\seq s$.
The pairs $(s_{j},t_{j})$ are called \emph{disagreement pairs.}
\end{defi}
    %
\begin{defi}
[$\jp D1$]\label{unif:def:opponents}
Given two terms $t$ and $s$,
let $\param[\seq x] t'$ and $\param[\seq y] s'$ 
be respective $\alpha$-equivalent terms such that
their parameters $\seq x$ and $\seq y$ are disjoint.
Then the disagreement pairs of $t'$ and $s'$ are called 
\emph{opponent pairs} in $t$ and $s$.
\end{defi}

\begin{lemma}[$\jp L{3\text{ (1)}}$]\label{lem:principal-lemma}Let $\varrho$ be a substitution
    and $X$, $Y$ be free variables such that $\varrho\left(X\,\seq s\right)=\varrho\left(Y\,\seq t\right)$
    for some term tuples $\seq s$ and $\seq t$. Then for every opponent
    pair $u$, $v$ in $\varrho X$ and $\varrho Y$
    (Definition \ref{unif:def:opponents}), the head of $u$ or $v$ is a parameter
    of $\varrho X$ or $\varrho Y$.
    \end{lemma}
  %

  In contrast to applied constants, applied variables should not be eagerly decomposed.
  For a constant $\cst{f}$, if $\cst{f}\,\seq s \unif \cst{f}\,\seq t$ has a unifier,
  that unifier must clearly also unify $s_{i} \unif t_{i}$ for each~$i$. 
  For a free variable $X$, a unifier of $X\,\seq s \unif X\,\seq t$ does not necessarily
  unify $s_{i} \unif t_{i}$.
  The concept of $\omega$-simplicity is
  a criterion on unifiers that captures some of the cases where eager decomposition is possible.
  Non-$\omega$-simplicity
  on the other hand is the main trigger of $\unifrulename{Iteration}$\textemdash the
  most explosive binding of our procedure.

  \begin{defi}
  [$\jp D2$]\label{unif:def:omega-simple}An occurrence of a parameter $x$ of term $t$
  in the body of $t$ is $\omega$\emph{-simple }if both
  \begin{enumerate}
  \item the arguments of $x$ are distinct and are exactly (the $\eta$-long
  forms of) all of the variables bound in the body of $t$, and
  \item this occurrence of $x$ is not in an argument of any parameter
  of $t$.
  \end{enumerate}
  \end{defi}
  This definition is slightly too restrictive for our purposes.
  It is unfortunate that condition~1 requires $x$ to be applied to
  \emph{all} instead of just some of the bound variables. The JP
  proof would probably work with such a relaxation, and the definition
  would then cover all cases where eager decomposition is possible. 
  However, to reuse the JP lemmas, we stick to the original
  notion of $\omega$-simplicity and introduce the following relaxation:
  \begin{defi}
  An occurrence of a parameter $x$ of term $t$ in the body of $t$
  is \emph{base-simple} if it is $\omega$-simple or both
  \begin{enumerate}
  \item $x$ is of base type, and
  \item this occurrence of $x$ is not in an argument of any parameter
  of $t$.
  \end{enumerate}
  \end{defi}
  \begin{lemma}\label{lem:base-simple->inj.}
  Let $s$ have parameters $\seq x$ and a subterm $x_{j}\,\seq v$ 
  where this occurrence of $x_{j}$ is base-simple.
  Then for any sequence $\seq t$ of (at least $j$) terms, the body of
  $t_{j}$ is a subterm of $s\,\seq t$ (after normalization) at the position
  of $x_{j}\,\seq v$ up to renaming of the parameters of $t_{j}$. 
  To compare positions of $s$ and $s\,\seq t$, ignore
  the parameter count mismatch.
  % ASK(ALEX), ASK(VISA): I am no expert, but isn't it unusual to use these
  % kind of instructions in the statement of the theorem? Maybe it would be good
  % to locally redefine the concept of position.
  \end{lemma}
  \begin{proof}
  Consider the process of $\beta$-normalizing $s\,\seq t$. After substituting
  terms $\seq t$ into the body of~$s$, a further reduction can only
  take place when some $t_{k}$ is an abstraction that gets arguments in
  $s$. The arguments $\seq v$ to the $x_{j}$ are distinct variables
  bound in the body of $s$. This follows easily from either case of the
  definition of base-simplicity. So $t_{j}$ is applied to the unmodified
  $\seq v$ after substituting
  terms $\seq t$ into the body of~$s$. Base-simplicity also implies
  that $t_{j}\,\seq v$ does not occur in an argument to another
  $t_{k}$. Hence only the reduction of $t_{j}\,\seq v$ itself affects this
  subterm. The variables $\seq v$ match the parameter count of $t_{j}$
  because we consider the $\eta$-long form of $t_{j}$; so $t_{j}\,\seq v$ reduces to the body
  of $t_{j}$ (modulo renaming). The position is obviously that of $x_{j}\,\seq v$.
  \end{proof}
  \begin{lemma}[$\jp C{4 strengthened}$]\label{lem:eq-if-omega-simple}Let $\varrho$ be a substitution and $X$ a free
    variable. If $\varrho\left(\param X\,\seq s\right)=\varrho\left(\param X\,\seq t\right)$
    and some occurrence of the $i^{\text{th}}$ parameter of $\varrho X$
    is base-simple, then $\varrho s_{i}=\varrho t_{i}$.
    \end{lemma}
    \begin{proof}
    By Lemma \ref{lem:base-simple->inj.}, $\varrho s_{i}$ occurs in
    $\varrho X\left(\varrho\seq s\right)$ at certain position that depends
    only on $\varrho X$. Similarly $\varrho t_{i}$ occurs in $\varrho X\left(\varrho\seq t\right)=\varrho X\left(\varrho\seq s\right)$
    at the same position, and hence $\varrho s_{i}=\varrho t_{i}$.
    \end{proof}
    We define more properties to determine which binding to apply to a given constraint.
    Roughly speaking, the simple comparison form will trigger identification bindings,
    projectivity will trigger Huet-style projections, and
    simple projectivity will trigger JP-style projections.
    \begin{defi}[$\jp D4$]
    \label{unif:def:simple-comparison-form}We say that $s$ and $t$ are in \emph{simple
    comparison form }if all $\omega$-simple heads of opponent pairs in $s$ and $t$ are distinct, and
    each opponent pair has an $\omega$-simple head.
    \end{defi}
    \begin{defi}[$\jp D5$]
    \label{unif:def:projective} A term $t$ is called \emph{projective }if the head
    of $t$ is a parameter of $t$. If the whole body is just
    the parameter, then $t$ is called \emph{simply projective.}
    \end{defi}
    A central part of the proof is to find a suitable measure for the
    remaining problem size. Showing that the measure is strictly decreasing
    and well-founded guarantees that the procedure finds 
    a suitable substitution in finitely many steps. We reuse the measure for remainder substitutions
    from JP \cite{jp-76-unif}, but embed it into a lexicographic measure
    to handle the decomposition steps and oracles of our procedure.
    \begin{defi}[$\jp D7$] \label{unif:def:weights}
    The \emph{free weight} of a term $t$ is the total number of occurrences of
    free variables and constants in $t$.
    The \emph{bound weight} of $t$ is the total number of occurrences 
    (excluding occurrences $\lambda x$) of bound variables in $t$,
    but with the particular exemption: if a prefix variable $u$ 
    has one or more $\omega$-simple occurrences in the body,
    then one such occurrence and its arguments are not counted.
    It does not matter which occurrence is not counted because
    in $\eta$-long form the bound weight of the arguments of an 
    $\omega$-simple variable is the same for all occurrences
    of that variable.
    \end{defi}
    \begin{defi}[$\jp D8$] \label{unif:def:ord}
    For multisets $E$ of unification constraints and
    substitutions $\rho$, our measure on pairs $(E, \rho)$
    is the lexicographic comparison of
    \begin{description}
      \item[A] the sum of the sizes of the terms in $\rho E$
      \item[B] the sum over the free weight of $\rho F$, 
        for all variables $F$ mapped by $\rho$
      \item[C] the sum over the bound weight of $\rho F$, 
      for all variables $F$ mapped by $\rho$
      \item[D] the sum over the number of parameters of $\rho F$, 
      for all variables $F$ mapped by $\rho$
    We denote the quadruple containing these numbers as $\ord(E,\rho)$.
    We denote the triple containing only the last three components of $\ord(E,\rho)$ as $\ord \rho$.
    %ASK(VISA): Why is ord(E,\rho) with parentheses and ord \rho without them?
    We write $<$ for the lexicographic comparison of these tuples.
    \end{defi}

    \looseness=-1
    The next six lemmas correspond to the bindings of our procedure and sufficient
    conditions for the binding to bring us closer to a given solution. 
    This is expressed as a decrease of the $\ord$ measure
    of the remainder. In each of these lemmas, 
    let $u$ be a term with a variable head $a$ and
    $v$ a term with an arbitrary head $b$.
    Let $\varrho$ be a unifier of $u$ and $v$.
    The conclusion, let us call it \textbf{\emph{C}},
    is always the same: there exists a binding $\delta$ applicable to the problem $ u\unif v$, and
    there exists a substitution $\varrho'$ such that $\ord\varrho'<\ord\varrho$
    and for all variables $X$ except the fresh variables introduced by the binding $\varrho X= \varrho'\, \delta X$.
    For most of these lemmas, we refer to JP \cite{jp-76-unif} for proofs.
    Although JP only claim $\varrho X= \varrho'\, \delta X$
    for variables $X$ mapped by $\varrho$,
    inspection of their proofs shows that the equality holds for 
    all $X$ except the fresh variables introduced by the binding.
    Moreover, some of our bindings have more preconditions, 
    yielding additional orthogonal hypotheses in our lemmas,
    which we address below.

    \begin{lemma}
    [$\jp L9$]\label{lem:elimination}If $a=b$ is not an elimination variable and $\varrho a$
    discards any of its parameters, then \textbf{C} by elimination.
    Moreover, for the elimination variable $G$ introduced by this elimination,
    $\varrho'G$ discards none of its parameters and has the body of $\varrho a$.
    \end{lemma}
    \begin{proof}
    Let $\varrho a=\param t$ and let $\left(x_{j_{k}}\right)_{k=1}^{i}$
    be the subsequence of $\seq x$ consisting of those variables which
    occur in the body $t$. It is a strict subsequence,\emph{ }since $\varrho a$
    is assumed to discard some parameter. Since the equal heads $a=b$ of the
    constraint $u\unif v$ are not elimination variables, elimination
    for $(j_{k})_{k=1}^{i}$ can be applied. Let $\delta=\{ a\mapsto\param G\,x_{j_{1}}\,\ldots\,x_{j_{i}}\} $
    be the corresponding binding. Define $\varrho'$ to be like $\varrho$
    except
    \[
    \varrho'a=a\quad\text{and}\quad\varrho'G=\param[x_{j_{1}}...x_{j_{i}}]t.
    \]
    Obviously $\varrho'G$ is a closed term and $\varrho X = \varrho'\,\delta X$
    holds for all $X \not= G$. Moreover $\ord\varrho'<\ord\varrho$, because
    free and bound weights stay the same ($\varrho a$ and $\varrho'G$ have the
    same body $t$) whereas the number of parameters strictly decreases. The definition
    of $\left(j_{k}\right)_{k=1}^{i}$ implies that $\varrho'G$ discards
    none of its parameters.
    \end{proof}

    %
    \begin{lemma}
    [$\jp L{10}$]\label{lem:iteration}
    Assume that
    there exists a parameter $x$ of $\varrho a$ such that
    $x$ has a non-$\omega$-simple (Definition \ref{unif:def:omega-simple}) 
    occurrence in $\varrho a$, which is not below another parameter, or
    such that $x$ has at least two $\omega$-simple occurrences in $\varrho a$.
    Moreover, 
    if $a = b$, to make iteration applicable,
    $a$~must not be an elimination variable, and
    $x$ must be of functional type.
    Then \textbf{C} is achieved by iteration.
    \end{lemma}
    %
    \begin{lemma}
    [$\jp L{11}$]\label{lem:jp-projection}
    \looseness=-1
    Assume that $a$ and $b$ are different free variables.
    If $\varrho a$ is simply
    projective (Definition \ref{unif:def:projective})
    and $a$ is not an identification variable, 
    then \textbf{C} by JP-style projection.
    \end{lemma}
    %\begin{proof}
    %This is Lemma 3.11 from \cite{jp-76-unif}, plus the assumption 
    %that $a$ is not an identification variable.
    %This extra assumption is necessary because we restrict 
    %the application of JP-style projection in this way.
    %\end{proof}
    %
    \begin{lemma}
    [$\jp L{12}$]\label{lem:jp-imitation}If $\varrho a$ is not projective and $b$ is rigid,
    then \textbf{C} by imitation.
    \end{lemma}
    %
    \begin{lemma}
    [$\jp L{13}$]\label{lem:identification}
    Let $a \not= b$.
    Assume that $\varrho a \not= a$ and $\varrho b \not= b$
    are in simple comparison form (Definition \ref{unif:def:simple-comparison-form}) and neither is
    projective. 
    Then \textbf{C} by identification.
    Moreover, $\varrho' H$ is not projective, where $H$ is the identification variable
    introduced by this application of the identification binding.
    \end{lemma}
    \begin{proof}
    This is JP's Lemma~3.13, plus the claim that $\varrho' H$ is not projective.
    Inspecting the proof of that lemma, it is obvious that $\varrho' H$ cannot be projective
    because $\varrho a$ and $\varrho b$ are not projective.
    \end{proof}
    %
    %Additionally, we need one more of these lemmas for the Huet-style projection binding:
    %
    \begin{lemma}
    \label{lem:huet-projection}
    Assume that $\varrho a$ is
    projective (Definition \ref{unif:def:projective}),
    $a$ is not an identification variable, and $b$ is rigid.
    Then \textbf{C} by Huet-style projection.
    \end{lemma}
    \begin{proof}
    Since $\varrho a$ is
    projective, we have $\varrho a = \lambda \overline{x}_n.\, x_k\,\overline{t}_m$
    for some $k$ and some terms $\overline{t}_m$.
    If $\varrho a$ is also simply projective,
    then $x_k$ must be non-functional
    and since Huet-style projection and JP-style projection coincide
    in that case, Lemma~\ref{lem:jp-projection} applies.
    Hence, in the following we may assume that $\varrho a$ is not simply projective,
    i.e., that $m>0$.

    Let $\delta$ be
    the Huet-style projection binding:
    \[ \delta = \{a \mapsto \lambda \overline{x}_n. \, x_i \, (F_1 \, \overline{x}_n) \, \ldots \, (F_m \, \overline{x}_n)\} \]
    for fresh variables $F_1,\dots,F_m$.
    This binding is applicable because $b$ is rigid.
    Let $\varrho'$ be the same as $\varrho$ except that we set $\varrho' a = a$
    and for each $1\leq j\leq m$ we set
    \[
      \varrho' F_j = \lambda \overline{x}_n. \, t_j
    \]
    It remains to show that $\ord\varrho'<\ord\varrho$.
    The free weight of $\varrho a$ is the same as 
    the sum of the free weights of $\varrho' F_j$ for $1\leq j\leq m$.
    Thus, the free weight is the same for $\varrho$ and $\varrho'$.
    The bound weight of $\varrho a$ however is exactly $1$ larger than
    the sum of the bound weights of $\varrho' F_j$ for $1\leq j\leq m$
    because of the additional occurrence of $x_k$ in $\varrho a$.
    The exemption for $\omega$-simple occurrences in the definition
    of the bound weight cannot be triggered by this occurrence of $x_k$
    because $m > 0$ and thus $x_k$ is not $\omega$-simple.
    It follows that $\ord\varrho'<\ord\varrho$.
    \end{proof}
    %
    We are now ready to prove the completeness theorem (Theorem \ref{thm:completeness}).
    \begin{proof}
    Let $E$ be a multiset of constraints
    and let $V$ be the supply of fresh variables provided to our procedure. 
    Let $\varrho$ be a unifier of $E$.
    We must show that there exists
    a derivation $\left(E,\makeop{id}\right)\longrightarrow^{*}\sigma$ and a
    substitution $\theta$ such that for all free variables $X\not\in V$, we have $\varrho X = \theta\sigma X$.

    Let $E_{0}=E$ and $\sigma_{0}=\makeop{id}$. 
    Let $\varrho_{0}=\tau\varrho$ for some renaming $\tau$, such that 
    every free variable occurring in $\varrho_{0}E_{0}$ does not occur in $E_{0}$ and is not contained in $V$.
    Then $\varrho_{0}$
    unifies $E_{0}$ because $\varrho$ unifies $E$ by assumption.
    Moreover, $\varrho_{0}=\varrho_{0}\sigma_{0}$. We proceed
    to inductively define $E_{j}$, $\sigma_{j}$ and $\varrho_{j}$
    until we reach some $j$ such that $E_j = \varnothing$.
    To guarantee well-foundedness, we ensure that the measure 
    $\ord\left(\varrho_{j},E_{j}\right)$ decreases with each step.
    We maintain the following invariants for all $j$:
    \begin{itemize}
    \item $(E_{j},\sigma_{j})\longrightarrow(E_{j+1},\sigma_{j+1})$;
    \item $\ord(\varrho_{j},E_{j}) > \ord(\varrho_{j+1},E_{j+1})$;
    \item $\varrho_{j}$ unifies $E_{j}$;
    \item $\varrho_{0} X = \varrho_{j}\sigma_{j} X$ for all free variables $X\not\in V$;
    \item every free variable occurring in $\varrho_{j}E_{j}$ does not occur in $E_{j}$ and is not contained in $V$;
    \item for every identification variable $X$, $\rho_j X$ is not projective; and
    \item for every elimination variable $X$, each parameter of $\rho_j X$ 
          has occurrences in $\rho_j X$,
          all of which are base-simple.
    \end{itemize}
    If $E_j \not= \varnothing$, let $u\unif v$ be the selected constraint $S(E_{j})$ in $E_{j}$.

    First assume that an oracle is able to find a CSU for the constraint $u\unif v$.
    Since $\varrho_{j}$ unifies $u$ and~$v$, by the definition of a CSU,
    the CSU discovered by the oracle 
    contains a
    unifier $\delta$ of $u$ and $v$ such that there exists $\varrho_{j+1}$
    and for all free variables $X$ except for the auxiliary variables of the CSU
    we have $\varrho_{j} X=\varrho_{j+1}\,\delta X$. Thus, an \unifrulename{OracleSucc}
    transition is applicable and yields the node 
    $(E_{j+1},\sigma_{j+1})=(\delta(E_{j}\setminus\{ u\unif v\} ),\delta\,\sigma_{j})$.
    %Since the auxiliary variables introduced by \unifrulename{OracleSucc} are fresh, they cannot occur in $E_{j}$. 
    Therefore we have a strict containment $\varrho_{j+1}E_{j+1}\subset\varrho_{j+1}\,\delta\,E_{j}=\varrho_{j}E_{j}$.
    This implies $\ord(E_{j+1},\varrho_{j+1})<\ord(E_{j},\varrho_{j})$.
    It also shows that the constraints $\varrho_{j+1}E_{j+1}$ are unified
    when $\varrho_{j}E_{j}$ are. 
    Since the auxiliary variables introduced by \unifrulename{OracleSucc} are fresh,
    they cannot occur in $E_j$ nor in $\sigma_{j} X$ for any $X\not\in V$. Hence,
    we have $\varrho_{0} X = \varrho_{j}\sigma_{j} X = \varrho_{j+1}\,\delta\,\sigma_{j} X =\varrho_{j+1}\sigma_{j+1} X$
    for all free variables $X\not\in V$.
    Any free variable occurring in $\varrho_{j+1}E_{j+1}$ cannot not occur in $E_{j+1}$ and is not contained in $V$
    because  $\varrho_{j+1}E_{j+1}\subset\varrho_{j}E_{j}$
    and the variables in $E_{j+1} = \delta(E_{j}\setminus\{ u\unif v\} )$
    are either variables already present in $E_{j}$ or fresh variables introduced by \unifrulename{OracleSucc}.
    New identification or elimination variables are not introduced;
    so their properties are preserved. Hence all invariants are preserved.

    Otherwise we proceed by a case distinction on the form of $u\unif v$.
    Typically, one of the Lemmas \ref{lem:elimination}\textendash{}\ref{lem:huet-projection}
    is going to apply. Any one of them gives substitutions $\varrho'$
    and $\delta$ with properties that let us define $E_{j+1}=\delta E_{j}$,
    $\sigma_{j+1}=\delta\,\sigma_{j}$ and $\varrho_{j+1}=\varrho'$.
    The problem size always strictly decreases, because 
    these lemmas imply
    $\varrho_{j+1}E_{j+1}=\varrho_{j+1}\delta E_{j}=\varrho_{j}E_{j}$
    and $\ord\varrho_{j+1}=\ord\varrho'<\ord\varrho_{j}$.
    Regarding the other invariants, the former equation guarantees that $\varrho_{j+1}$
    unifies $E_{j+1}$, and 
    $\varrho_{0} X = \varrho_{j}\,\sigma_{j} X = \varrho_{j+1}\,\delta\,\sigma_{j} X =\varrho_{j+1}\,\sigma_{j+1} X$
    for all $X \not\in V$ because the fresh variables introduced by the binding cannot occur in $\sigma_{j} X$ for any $X\not\in V$. The conditions on variables must be checked
    separately when new ones are introduced. Let $a$ be
    the head of $u=\param a\,\seq u$ and $b$ be the head of $v = \param b\,\seq v$.
    Consider the following cases:

    \begin{description}
    \item[$u$ and $v$ have the same head symbol $a=b$] \hfill
    \begin{enumerate}[beginpenalty=1000]
    \item Suppose that $\varrho_{j}a$ has a parameter with non-base-simple
    occurrence. By one of the induction invariants, $a$ is not an elimination
    variable. 
    Among all non-base-simple occurrences of parameters in $\varrho_{j}a$, 
    choose the leftmost one, which we call $x$.
    This occurrence of $x$ cannot be below another parameter, 
    because having $x$ occur in one of its arguments would make
    that other parameter non-base-simple, contradicting the occurrence of $x$ being leftmost.
    Thus $x$ is neither base-simple nor below another parameter; so $x$ is of functional type.
    Moreover, non-base-simplicity implies non-$\omega$-simplicity.
    Hence, we can apply Lemma~\ref{lem:iteration} (iteration).

    \item Otherwise suppose that $\varrho_{j}a$ discards some of its parameters.
    By one of the induction invariants, $\varrho_{j}a$ is not an elimination variable.
    Hence Lemma \ref{lem:elimination} (elimination) applies. The newly introduced elimination
    variable $G$ satisfies the required invariants, because Lemma \ref{lem:elimination}
    guarantees that $\varrho_{j+1}G$ uses its parameters and shares the
    body with $\varrho_{j}a$ which by assumption of this case contains
    only base-simple occurrences.

    \item Otherwise every parameter of $\varrho_{j}a$ has occurrences and
    all of them are base-simple. We are going to show that $\unifrulename{Decompose}$
    is a valid transition and decreases $\varrho_{j}E_{j}$. By Lemma~\ref{lem:eq-if-omega-simple}
    we conclude from $\varrho_{j}u=\varrho_{j}v$ that $\varrho_{j}u_{i}=\varrho_{j}v_{i}$
    for every $i$. Hence the new constraints 
    $E_{j+1}=E_{j}\setminus\left\{ u\unif v\right\} \cup\set{u_{i}\unif v_{i}}{\text{for all }i}$
    after $\unifrulename{Decompose}$ are unified by $\varrho_{j}$.
    This allows us to define $\varrho_{j+1}=\varrho_{j}$ and
    $\sigma_{j+1}=\sigma_{j}$.
    To check that $\varrho_{j+1}E_{j+1}=\varrho_{j}E_{j+1}$ is smaller
    than $\varrho_{j}E_{j}$ it suffices to check that constraints $\varrho_{j}u_{i}\unif\varrho_{j}v_{i}$
    together are smaller than $\varrho_{j}u\unif\varrho_{j}v$. Since
    all parameters of $\varrho_{j}a$ have base-simple occurrences, $\varrho_{j}u_{i}$
    is a subterm of $\varrho_{j}u=\param\varrho_{j}a\,\left(\varrho_{j}\seq u\right)$
    by Lemma \ref{lem:base-simple->inj.}. Similarly for $\varrho_{j}v$.
    It follows that $\varrho_{j+1}E_{j+1}$ is smaller
    than $\varrho_{j}E_{j}$.
    Since $\varrho_{j+1}=\varrho_{j}$ and
    $\sigma_{j+1}=\sigma_{j}$, the other
    invariants are obviously preserved.

    \end{enumerate}

    \item[\mbox{$u \unif v$} is a flex-flex pair with different heads] \hfill
    \begin{enumerate}
      \setcounter{enumi}{4}
      \item First, suppose that
      $\varrho_j a$ or $\varrho_j b$ is simply projective
      (Definition \ref{unif:def:projective}). 
      By the induction hypothesis, the simply projective head 
      cannot be an identification variable.
      Thus Lemma \ref{lem:jp-projection} (JP-style projection) applies.

      \item \looseness=-1 
      Otherwise suppose that 
      $\varrho_j a$ is projective but not simply.
      Then the head of $\varrho_j a$ is some parameter $x_{k}$. But
      this occurrence cannot be $\omega$-simple because it has arguments
      which cannot be bound above the head $x_{k}$. Thus Lemma \ref{lem:iteration}
      (iteration) applies.
      If $\varrho_j b$ is projective but not simply, the same argument applies.

      \item Otherwise suppose that $\varrho_j a$, $\varrho_j b$ are in simple
      comparison form (Definition \ref{unif:def:simple-comparison-form}).
      By one of the the induction invariants,
      the free variables occurring in $\varrho_{j}E_{j}$ do not occur in $E_{j}$.
      Thus $\varrho_{j} a \not= a$ and $\varrho_{j} b \not= b$.
      Then Lemma \ref{lem:identification} (identification) applies.

      \item Otherwise $\varrho_j a$, $\varrho_j b$ are not in simple comparison
      form. By Lemma \ref{lem:principal-lemma} and by the definition of simple comparison form,
      there is some opponent pair $x_{k}\,\seq r$, $ b$
      in $\varrho_j a$ and $\varrho_j b$
      (after possibly swapping $u$ and $v$) where either the occurrence
      of $x_{k}$ is not $\omega$-simple (Definition~\ref{unif:def:omega-simple}) or else
      $x_{k}$ has another $\omega$-simple occurrence in the body of $\varrho_j a$.
      Then Lemma \ref{lem:iteration} 
      (iteration) applies.
    \end{enumerate}

    \looseness=-1
    \item[$u \unif v$ is a flex-rigid pair]
    Without loss of generality, assume that $a$ is flex and $b$ is rigid.\hfill

    \begin{enumerate}
    \setcounter{enumi}{8}
    \item Suppose first that $\varrho_j a$ is projective.
    By one of the induction invariants, $a$ cannot be an identification variable.
    Thus Lemma \ref{lem:huet-projection} (Huet-style projection) applies.

    \item Otherwise $\varrho_j a$ is not projective. 
    The head of $\varrho_j a$ must be $b$ because $b$ is rigid, and $\varrho_j$ unifies $u$ and $v$. 
    Since $\varrho_j a$ is not projective, that means that $b$ is not a bound variable. 
    Therefore, $b$ must be a constant.
    Then Lemma \ref{lem:jp-imitation} (imitation) applies.
    \end{enumerate}

    We have now constructed a run 
    $(E_{0},\sigma_{0})\longrightarrow(E_{1},\sigma_{1})\longrightarrow(E_{2},\sigma_{2})\longrightarrow\cdots$
    of the procedure. 
    This run cannot be infinite because because the measure $\ord(E_j,\varrho_{j})$
    strictly decreases as $j$ increases.
    Hence, at some point we reach a $j$ such
    that $E_{j}=\varnothing$ and $\varrho_{0} X = \varrho_{j}\,\sigma_{j} X$ for all $X \not\in V$.
    Therefore, 
    $(E,\makeop{id})\longrightarrow^*(\varnothing,\sigma_{j})\longrightarrow\sigma_j$,
    and $\rho X = \tau^{-1}\,\varrho_{j}\,\sigma_{j} X$ for all $X \not\in V$,
    completing the proof. \qedhere
    \end{description}
    \end{proof}

\pagebreak[2]    
\section{A New Decidable Fragment}
\label{sec:unif:solid-oracle}

We discovered a new fragment that admits a finite CSU and a simple oracle. The
oracle is based on work by Prehofer and the PT procedure \cite{cp-95-unifphd},
an adaptation of the preunification procedure by Snyder and Gallier
\cite{sg-89-unif} (which itself is an adaptation of Huet's procedure). PT
transforms an initial multiset of constraints $E_0$ by applying bindings
$\varrho$. If there is a sequence
$E_0\ptarrow{\varrho_1}\cdots\ptarrow{\varrho_n}E_n$ such that $E_n$ has only
flex-flex constraints, we say that PT produces a preunifier $\sigma =
\varrho_n\ldots\varrho_1$ with constraints $E_n$. A sequence fails if
$E_n=\bot$. As in the previous section, we consider all terms to be
$\alpha\beta\eta$-equivalence classes with the $\eta$-long $\beta$-reduced form
as their canonical representative. Unlike previously, in this section we view
unification constraints $s \unif t$ as ordered pairs.

The following rules, however, are stated modulo orientation. The PT transition
rules, adapted for our presentation style, are as follows:
